\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-20 paper.tex
Drop some changes that will be nuked

\linenumbers
\section{Introduction}
\label{sec-intro}
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
% and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
% TODO finish this sentence with "and more reasons why ARGs are cool"
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs.

% TODO need to mention ARG-as-a-data-structure to clarify.
% What is the problem?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG  has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the outcome of such a stochastic process or a
genetic genealogy inferred from a sample of genome sequences.
However, the distinction between process and data structure is not clear cut
and subfields use the term
differently (see Appendix~\ref{sec-arg-history}).

% JK: this para need a major update, it's quite weak right now.
The word ``ARG'' is now widely used to refer to genetic genealogies in
general~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage means there is a lack of consensus on the
exact object being defined, which causes problems, for example,
when evaluating inferred ARGs, and even when defining
what the criteria for success should be.
It also means there is no shared format for interchanging ARG data between
programs, resulting in the assorted negative consequences with which
geneticists are only too familiar~\citep{excoffier2006computer,gower2022demes}.
Perhaps most importantly, definitions linked to
the original stochastic process lead to an
overly narrow interpretation of ARGs, which is significantly
out of step with modern methods and large-scale datasets, and
can lead to missed opportunities in their application and usage.
Such problems go much deeper than pedantic discussion over terminology;
they directly inform what we should be attempting to estimate, how we
assess success, and the expected limitations on the precision
and accuracy of our inferences.
% Thanks for the phrase Gideon and Alex!
If not addressed, these problems will
impede progress in applying genetic genealogies to understand the biological world.

% Evidence of confusion to cite above in "problem" para?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/334
% The output of these programs is certainly not a collection of
% \emph{unrelated} local trees, as implied by,
% e.g.,
% \citet{hejase2020summary} when they state that \tsinfer\
% ``does not explicitly infer an ARG but rather a sequence of local
% gene trees''.
% More cites? Not worth harping on about now?
% https://github.com/tskit-dev/what-is-an-arg-paper/issues/38

@@

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we define a genome ARG (gARG) and contrast it with event ARGs (eARGs).
% Retrospective vs prospective
% TODO: This needs revision - see also https://github.com/tskit-dev/what-is-an-arg-paper/pull/380
We continue with a description of the usually held retrospective view of ARGs,
but note that ARGs can also be viewed as prospective, particularly in forwards-time
simulations. In the context of forwards-time simulations, we mention the
simplification operation, which removes ancestral genome material that was not
inherited.
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

\section{Genome ARGs}
\label{sec-gARG}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes and their coordinate spaces, and hope that the distinction
will be clear from the context.
We are not concerned here with mutational processes or observed sequences,
but consider only processes of inheritance,
following the standard practice in coalescent theory.

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, and $E$ be the set of edges.
We will usually define a set of nodes $S \subseteq N$
representing the sampled genomes
(but see Section~\ref{sec-retro-pro} for some subtlety around this point).
Each element of $E$ is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is ``annotated'' with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes can respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

% TODO add a few sentences here on why defining nodes concretely as
% integers has significant advantages.
% 1) leads to direct link between the ARG and local trees in terms
%    of a well-defined combinatorial object (local trees section)
% 2) Definitions can be implemented directly in an efficient way
%    (computational efficiency section.
In many settings, nodes are dated, i.e.\
each node $u\in N$ is associated with a time $\tau_u$.
However, the topological relationships
defined by the directed graph structure is sufficient for
many applications, and we will assume that node dates are not
known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
Following the above definitions, corresponding data encoding for
this gARG is shown in Tab.~\ref{tab-gARG-data}.

These definitions follow naturally from the outcome of events
associated with the inheritance of DNA between generations.
% FIXME this sentance doesn't follow well from the first, needs a
% different bridge before listing important properties.
% Maybe the "listicle" format here isn't right anyway.
As such,
there are some important consequences to these definitions.
Firstly, there is
no limitation on a mating system, ploidy, or age structure in a gARG.
By making the
basic unit an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance can be directly expressed, such as
gene conversions,
multiple simultaneous recombinations,
and so on.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
% TODO, should probably drop this point.
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity (number of child nodes), including ``unary''
nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
% TODO: check above if we want to refer to sec-ARG-simplification or some other!
parental nodes from which this child has inherited genome segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.

@@

\section{Event ARGs}
\label{sec-eARG}
% FIXME - this intro doesn't work at all now - we haven't defined what an
% eARG *is* before discussing it.
% Maybe best to skip this entirely, and make the quoted definitions with
% a thin intro the way to start the section.
Genome ARGs are defined by
describing the details of genetic inheritance
between ancestor and descendant genomes.
In contrast, ARGs are classically defined
by describing the \emph{events} in the history of a sample of genomes.
The difference between gARGs and the prevalent views of eARGs
% what defines an ARG
are subtle and difficult to distinguish.
It is therefore useful to first survey some representative
recent definitions of
an ARG from the literature, to allow us to capture their common features
(see Appendix~\ref{sec-arg-history} for a full review).

\citet{brandt2021evaluation} follow the original Griffiths definition of
an ARG by coupling it directly to a generative process, stating
that the ``full ancestral recombination graph (ARG) is a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination.''
\citet{shipilina2023origin} emphasise the true existence of the
corresponding events (rather than the generative process) by saying that
``[t]he ARG describes the complete
ancestry of a sample of genomes through a series of real coalescence and
recombination events.''
\cite{mathieson2020ancestry}
define an ARG as ``a subset of the pedigree...
contain[ing] only those edges along which inherited segments of DNA have been
transmitted, and only those nodes corresponding to ancestors in which there was
a recombination or coalescence event.''
% In contrast, ~\citet{gusfield2014recombinatorics} defines an ARG as [TODO]
% Papers that do not discuss how positions work:
% - Mathieson and Scally: no discussion of how it works concretely (but quite
%   close to gARG, in some ways)
% TODO list out some of these as examples of where people don't bother
% getting into this important detail?

% NOTE: important to emphasise TWO things here, events and vagueness
% actual encoding mechanism (i.e. how trees fit into the ARG)
A common feature of these definitions is that they are expressed
in terms of evolutionary \emph{events}, and so for the purposes of
discussion we can label
this prevailing interpretation as an ``event ARG'', or eARG.
(Many definitions also emphasise
the necessity of exhaustively capturing \emph{all} events, which we
examine in later sections.)
Another common feature of these definitions is that they do not explicitly
discuss the mechanism by which local trees vary along the genome.
The topology of an eARG defines the common ancestor and
recombination events that occurred,
but the graph structure alone is not sufficient to define the local trees,
or equivalently, how ``ancestral material'' flows along
the edges of the ARG (see Section~\ref{sec-eARG-to-gARG}).
% TODO: check above if we want to refer to sec-eARG-to-gARG or some other!
This is usually assumed to follow the mechanism described by
Griffiths and colleagues in which we have
two different ``types'' of node in the graph:
common ancestor nodes in which the inbound lineages are merged into a
single ancestral lineage with one parent, and recombination
nodes in which a single lineage is split into two independent
ancestral lineages.
Recombination nodes are then
annotated with
the corresponding crossover breakpoint~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event,
% Trying to emphasise here that this is just an example data encoding.
% I can imagine
% people nitpicking about how this specific way of doing it is bad
% for reason X.
along with an example encoding of the data.
There is a single recombination event
at node \noderef{d} with a breakpoint at position $x$. We
assume that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
As shown in Fig.~\ref{fig-event-arg}B,
the local trees are embedded in the graph, and can be reconstructed
in essentially the same manner as outlined in the previous section for
gARGs. The only difference is that decisions about which outbound
edge to follow during the rootward traversal are taken at recombination
nodes, following
either the common convention that the order of parents
is significant~\citep[e.g.][]{griffiths1991two},
or by incorporating additional information to
distinguish left and right parents
into the
encoding~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}.
% JK: not sure what you're getting at here Gregor? The ordering matters
% because you have to write each event down precisely.
% Gregor: Importantly, the order of events matters to WHAT!?

@@

\node [anchor=north west] at (\x1,6) {A};
\node [anchor=north west] at (\x2,6) {B};
\node [anchor=north west] at (\xt,6) {C};

@@

%%% (C) Encoding
\node [nodelabel,anchor=north west] at ($(\xt,5)$) {
\begin{tabular}{c|c|l}
% \multicolumn{2}{c}{Breakpoints}\\
Node & Breakpoint & Parents\\
\hline
$\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
$\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
$\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
$\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
$\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
$\noderef{g}$ & $\varnothing$ & []\\
\end{tabular}};

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
for a discussion of this important point).
(C) A concrete encoding of the information.
Recombination nodes have a non-null breakpoint
and two parents.
}
\end{figure}

% TODO reference forward to the cell ARG appendix. Ultimately
% every event corresponds to a particular genome in a particular
% individual or cell.
Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that in an eARG we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
whereas in a gARG we store \emph{inheritance intervals} on edges.
A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe genetic inheritance events other
than simple crossovers, such as multiple recombinations and
gene conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.

>>>>

\linenumbers
\section{Introduction}
\label{sec-intro}
% Para 1: ARGs are cool!
Estimating the genetic genealogy of a set of genome sequences
under the influence of recombination,
usually known as an Ancestral Recombination Graph (ARG), is a long-standing
goal in genetics.
Broadly speaking, an ARG describes the different paths of genetic inheritance
caused by recombination, and encodes a sequence of correlated genealogical
trees along the genome.
Recent breakthroughs
in large-scale inference
methods~\citep{rasmussen2014genome,kelleher2019inferring,speidel2019method,
schaefer2021ancestral,wohns2022unified,zhang2023biobank,zhan2023towards}
% and data representation~\citep{kelleher2016efficient,kelleher2018efficient}
have raised the realistic prospect of ARG-based analysis becoming a standard part
of the population and statistical genetics toolkit~\citep{hejase2020summary}.
Applications using inferred ARGs as input have begun to
appear~\citep{osmond2021estimating,
fan2022genealogical,
hejase2022deep,
guo2022recombination,
zhang2023biobank,
nowbandegani2023extremely,
ignatieva2023distribution,
fan2023likelihood}
and many more are sure to
follow~\citep{harris2019database,harris2023using}.
See \citet{lewanski2023era} for a biologically oriented introduction to ARGs
and further discussion on the potential impacts.

% Para 2: What are they, though?
Although it is widely accepted that ARGs are important, there is significant
confusion about what, precisely, an ARG \emph{is}.
Originally, ARGs were defined by Griffiths and colleagues as an alternative
formulation of the coalescent with recombination~\citep{hudson1983properties},
where the stochastic process of coalescence and recombination
among ancestral lineages is formalised as a
% Note: please don't add "mathematical graph". If a reader doesn't know what a
% graph is they're not going to understand this paper.
graph~\citep{griffiths1991two,ethier1990two,griffiths1996ancestral,griffiths1997ancestral}.
Subsequently, an ARG  has come to be thought of as a data
structure~\citep{minichiello2006mapping}, i.e.\ describing
the \emph{outcome} of such a random process or an
% Note on earlier version: introducing a different term for ARG here is just confusing
inference from a sample of genome sequences.
The distinction between stochastic process
and data structure is not clear cut, however,
and subfields use the term
differently (see Appendix~\ref{sec-arg-history}).
% pedantically, ARG is not really a "word"
The term ``ARG'' is now most often used in quite a general
sense~\citep[e.g.][]{mathieson2020ancestry,hejase2020summary,
schaefer2021ancestral,harris2023using,zhang2023biobank,fan2023likelihood},
as a data structure
encompassing the varied outputs from modern simulation and
inference methods~\citep{rasmussen2014genome, palamara2016argon, haller2018tree,
kelleher2019inferring, speidel2019method, baumdicker2021efficient, zhang2023biobank}.
This broad usage seems a useful generalisation,
but is not compatible with the usual concrete definition of an ARG.
This mismatch has led to a range of problems that, if unchecked,
may impede progress in applying ARGs to understand the biological world.

% Para 3: what is the problem?
% FIXME current version isn't great, but it's better than previous effort.
[FIXME this paragrph needs a substantial update, need to reread the
eARG and gARG sections again to be clear what we say we're doing.
We should focus on the concrete interchange aspects rather than
coalescenty type things.]
The mismatch between the classical Griffiths definition of an ARG and the
modern interpretation is the source of substantial confusion,
and, we argue, a significant limiting factor for the field.
% actual outputs of these modern methods is a central issue.
As this definition is essentially a reformulation of the
coalescent with recombination (henceforth ``the coalescent''; we
will assume the incorporation of recombination for the purposes of
this article) in graph-theoretic form,
the distinction between the simplifying assumptions of the model
and the necessary requirements for a data structure are not clear.
The consequences of conflating model with data structure
are subtle, varied, and deep.
% TODO rephrase, not great currentyl
We argue that the prevailing definition of an ARG
imposes unneccesary limitations, and that with a broader definition
we can reassess what we should be attempting to estimate,
how we measure success, and the expected limitations on the precision
and accuracy of our inferences.

@@

% TODO: Provide a map of the paper's sections and how they tie together in support of gARG as a way forward.
% Here is a rough draft - it would be good to state the most important message (in one sentence or so) from each of the sections to help with frontloading key messages
% Genome ARGs &
% Event ARGs
In the following, we first define the genome ARG encoding
and contrast it with the event ARG approach.
We then discuss the important concept of ancestral material in
Section~\ref{sec-ancestral-material}
and [DO SOME STUFF].
% Converting and eARG to gARG &
% ARGs and local trees &
% Locally unary nodes &
% Levels of simplification
This simplification is closely related to operations needed to convert eARG and gARG,
which opens a rich set of details about ARGs and local trees, local unary nodes, and
levels of simplification.
% Precision of recombination information
We then highlight the important difference between the required level of precision
about recombination information for eARGs compared to gARGs.
% Computational efficiency and data interchange
The way this recombination information is stored in eARGs and gARGs has important
implications for computational efficiency, which is critical if we want to use ARGs
for modern genomic datasets.
% Discussion
We wrap with a discussion highlighting that WHAT-IS-THE-KEY-SINGLE-MESSAGE-NOT-COVERED-ABOVE?
Finally, the literature on ARGs and associated aspects is vast, particularly
due to different definitions and implementations. We cover some important aspects
in appendices, including
a brief history of ancestral graphs (Appendix~\ref{sec-arg-history}),
description of the big and little ARGs (Appendix~\ref{sec-big-and-little-arg}),
survey of ARG inference methods (Appendix~\ref{sec-survey-arg-infer}), and
ARGs at an individual vs cell lineage level (Appendix~\ref{sec-cell-lineages-and-args}).

\section{Genome ARGs}
\label{sec-gARG}
We define a genome as the complete set of genetic material that a child
inherits from one parent (i.e.\ haploids have one genome, diploids two, etc.).
We will also use the term ``genome'' in its
more common sense of ``the genome'' of a species, denoting the full set
of chromosomes, and hope that the distinction
will be clear from the context.
We are not concerned here with mutational processes or observed sequences,
but consider only processes of inheritance,
following the standard practice in coalescent theory.
We also do not consider structural variation, and assume that all
samples and ancestors share the same genome coordinate space.

A genome ARG (gARG) is a directed acyclic graph in which nodes represent
haploid genomes and edges represent
genetic inheritance between an ancestor and a descendant.
The topology of a gARG specifies that genetic inheritance
occurred between particular
ancestors and descendants, but the graph connectivity
does not tell us which \emph{parts} of their genomes were inherited.
In order to capture the effects of recombination
(and other processes that lead to the partial genome inheritance)
 we ``annotate'' the edges with the genome
coordinates over which inheritance occurs.
We can define a gARG formally as follows.
Let $N \subset \mathbb{N}$ be the set of nodes representing
the genomes in the gARG, $S \subseteq N$ be the set of sampled genomes.
% (but see Section~\ref{sec-retro-pro} for some subtlety around this point).
Then, $E$ is the set of edges, where each element
is a tuple $(c, p, I)$ such that $c, p \in N$ are the child and
parent nodes and $I$ is the set of disjoint genomic intervals
over which genome $c$ inherits from $p$.
Thus, each topological connection between
a parent and child node in the graph is annotated with a set of
inheritance intervals $I$.
Here, the terms parent and child are used in graph sense;
these nodes respectively represent ancestor and descendant genomes,
which can be separated by multiple generations.
We will use these two sets of terms interchangeably.

In many settings, nodes are dated, i.e.\ each
node $u\in N$ is associated with a time $\tau_u$.
The topological relationships
defined by the directed graph structure, however,
 is sufficient for many applications,
and we will assume that node dates are not known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.
Concretely defining nodes as integers in this way has some
significant advantages. In particular, it provides a direct link
between an ARG and its local trees in terms of a well-defined
combinatorial object (Section~\ref{sec-ARG-and-local-trees}),
as well as enabling straightforward connection between
definitions and practical implementation (Section~\ref{sec-efficiency}).

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationships between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations (A), the corresponding gARG (B), and finally the corresponding
sequence of local trees  along the
genome (C).
% Jerome: this feels unecessary and "genealogy" is often interpreted as
% "pedigree" so needs to be used very carefully.
% The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
% which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
See Section~\ref{sec-ARG-and-local-trees} for further details on
how the local trees are embedded in an ARG.
% Following the above definitions, corresponding data encoding for
% this gARG is shown in Tab.~\ref{tab-gARG-data}.

% These definitions follow naturally from the outcome of events
% associated with the inheritance of DNA between generations.
% % FIXME this sentance doesn't follow well from the first, needs a
% % different bridge before listing important properties.
% % Maybe the "listicle" format here isn't right anyway.
% As such,
% there are some important consequences to these definitions.
% Firstly, there is
% no limitation on a mating system, ploidy, or age structure in a gARG.
% By making the
% basic unit an individual's genomes,
% any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
% species) can be accommodated.
% Secondly, because we record the intervals of inheritance
% between genomes,
% arbitrary patterns of genetic inheritance can be directly expressed, such as
% gene conversions,
% multiple simultaneous recombinations,
% and so on.
% % Wilder: give forward ref to section on efficiency here, mention tree sequence
% % TODO, should probably drop this point.
% Thirdly, local trees that we generate for a particular position on the
% genome can have nodes of any arity (number of child nodes), including ``unary''
% nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
% % TODO should rephrase this from parent/child to ancestor/descendant?
% Fourth, recombination is modelled in terms of
% its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
% Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
% Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
% % TODO: check above if we want to refer to sec-ARG-simplification or some other!
% parental nodes from which this child has inherited genome segments.
% % TODO rephrase the above, it'll confuse people. Just need to reiterate the
% % points about events, and maybe lead into the next section
% Finally, gARGs are not defined in terms of the
% coalescent with recombination (or any other) stochastic process.
% Thus, the gARG encoding is focused on faithfully representing the
% \emph{outcome} of the processes of genetic inheritance
% with recombination,
% and is agnostic to the details of those processes.

@@

\section{Event ARGs}
\label{sec-eARG}
% % I feel like sophisticated readers would at this point be saying,
% % ah, yeah, that's basically what an ARG is just in different words;
% % what's the point of this paper?
The genome ARG framework defined in the previous section is
in many ways a clarification of existing treatments
\cite[e.g.][]{mathieson2020ancestry,shipilina2023origin},
adding concrete details to the mechanisms by which
differential inheritance of genetic material are modelled.
% This is definitely needed for the Nicks, who are always thinking
% about the deeper processes and are quite happy being loosey-goosey
% with representation details
Note that here, and throughout, we are not questioning the form of the
actual ancestral processes that occur in nature, but rather how we
\emph{represent} such realised histories in a practical sense.
% Again, need to remind the reader that there's an actual point
% to all this getting stuck into the details.
These practical details,
as demonstrated through the remainder of the article, have important
consequences not only for how methods exchange information about
simulated and inferred ARGs, but more fundamentally in
how we set our goals for inference and evaluate the success of results.

% section is in many ways simply a clarification of existing definitions,
% it does provide substantial advantages (as shown in later sections).
% To allow us to concretely discuss and characterise the prevailing
% definition of an ARG, we begin by surveying some
% recent definitions from the literature.
% \citet{brandt2021evaluation} follow the original Griffiths definition of
% an ARG by coupling it directly to a generative process, stating
% that the ``full ancestral recombination graph (ARG) is a structure that encodes all
% coalescence and recombination events resulting from the stochastic process of
% the coalescent with recombination.''
% \citet{shipilina2023origin} emphasise the true existence of the
% corresponding events (rather than the generative process) by saying that
% ``[t]he ARG describes the complete
% ancestry of a sample of genomes through a series of real coalescence and
% recombination events.''
% \cite{mathieson2020ancestry}
% define an ARG as ``a subset of the pedigree...
% contain[ing] only those edges along which inherited segments of DNA have been
% transmitted, and only those nodes corresponding to ancestors in which there was
% a recombination or coalescence event.''

% The prevailing view, captured in these representative definitions, is that an
% ARG is composed of \emph{events}, and so we will loosely label this approach as
% ``event ARGs'', or ``eARGs''. Whether we view nodes in an ARG as events or
% genomes is essentially a philosophical question and in most senses the two are
% equivalent (see Appendix~\ref{sec-cell-lineages-and-args}, however, for a discussion
% of some subtle issues around \emph{which} genome corresponds to events
% of various types).
% How we represent and reason about the differential inheritance of
% genetic matierial \emph{between} nodes in an ARG, however, is of fundamental
% importance, and the central theme of this article.
% % Need to make this point strongly for the Nicks. Not sure this is the
% % right place, though. Maybe it's good to remind them again here though...
% Note that here, and throughout, we are not questioning the form of the
% actual ancestral processes that occur in nature, but rather how we
% \emph{represent} such realised histories in a practical sense.

@@

\node [anchor=north west] at (\x1,6) {A};
\node [anchor=north west] at (\x2,6) {B};
% \node [anchor=north west] at (\xt,6) {C};

@@

% %%% (C) Encoding
% \node [nodelabel,anchor=north west] at ($(\xt,5)$) {
% \begin{tabular}{c|c|l}
% % \multicolumn{2}{c}{Breakpoints}\\
% Node & Breakpoint & Parents\\
% \hline
% $\noderef{a}$ & $\varnothing$ & [\noderef{e}]\\
% $\noderef{b}$ & $\varnothing$ & [\noderef{d}] \\
% $\noderef{c}$ & $\varnothing$ & [\noderef{f}]\\
% $\noderef{d}$ & $x$ & [\noderef{e}, \noderef{f}] \\
% $\noderef{e}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{f}$ & $\varnothing$ & [\noderef{g}]\\
% $\noderef{g}$ & $\varnothing$ & []\\
% \end{tabular}};

\end{tikzpicture}
\caption{\label{fig-event-arg}
A classical event ARG. (A) Standard graph depiction with
breakpoint $x$ associated with the recombination node \noderef{d}.
Nodes \noderef{e}, \noderef{f} and \noderef{g} are common ancestor events.
(B) Corresponding local trees to the left and right of breakpoint $x$
(note these are shown in the conventional form in which only coalescences
within the local tree are included; see Section~\ref{sec-ARG-and-local-trees}
% TODO: check above if we want to refer to sec-ARG-and-local-trees or some other!
for a discussion of this important point).
}
\end{figure}

% The prevailing view is that ARGs are composed of events, and so we
% characterise this view
% Besides viewing ARGs as being composed of events, another
% feature of the definitions given above is that they do not
% explicitly describe
% % TODO check this! Do they talk about how trees are embedded?
% the details of how such events are represented.
Where the concrete details of an ARG data structure
(the encoding) are
% TODO more?
specified~\citep[e.g.][]{
rasmussen2014genome,gusfield2014recombinatorics,ignatieva2021kwarg}
they follow the approach described by Griffiths and colleagues.
Because it is focused on recording the details of events,
we will refer to this Griffiths encoding as an ``Event ARG'', or eARG.
In an eARG, we have two different types of node in the graph
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoint, and these breakpoints
are used to reconstruct the local trees.
This is done by tracing rootwards through the graph from the samples,
taking decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
We assume here that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.
